Adaptive Algorithms Employing Continuous Linear Functionals

### 2018-08-03 AutoAlgorithmLinearFunctionalDataVerC.tex
John
revising intro

Let $\ch$ be any subset of $\cf$, and let $\ca(\ch)$ denote the set of deterministic algorithms that successfully approximate the solution to some error tolerance for all inputs in $\ch$:
\begin{multline} \label{adapErrCrit}
\ca(\ch) : = \left\{ \text{algorithms } A:\ch \times (0,\infty) \rightarrow \cg : 
\right . \\ \left .
\bignorm[\cg]{S(f) - A(f,\varepsilon)} \le \varepsilon \ \forall f \in \ch, \ \varepsilon > 0 
\right \}.
\end{multline}
Algorithms in $\ca(\ch)$ are allowed to sample adaptively any bounded, linear functionals
of the input function $f$.  The definition of $\ch$ can be used in constructing algorithms in $\ca(\ch)$, but no other a priori knowledge about the input functions is available.

@@

\begin{theorem} \label{thm:lowbdcomp}
A lower bound on the complexity of the linear problem defined in \eqref{probDef} is
\begin{subequations} \label{compbd}
\begin{align}
 \label{compbdA}
\comp(\ca(\cc),\varepsilon,\rho) &\ge n_{j^*},  \\
\label{compbdB}
\text{where} \qquad j^* &= \max \left \{ j \in \naturals : \sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}}   < \frac{\rho(a-1)}{2a \varepsilon}
\right \}.
\end{align}
\end{subequations}
\end{theorem}

@@

The first test function $f \in \cc$ is defined in terms of its series coefficients:
\begin{align}
\nonumber
\hf_i &:= \begin{cases}
\displaystyle
\frac{c b^{k-j}}{\lambda_{n_{k}}},  & i =  n_{k}, \ k = 1, \ldots, j,
\\
0, & \text{otherwise}
\end{cases}
\\
\label{cdef}
c &:= \frac{\rho (a+1)}{2a} \left[\sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}} \right]^{-1}.
\end{align}
It can be verified that the test function lies in a ball of radius $\rho$ and also in the cone $\cc$:
\begin{align}
\label{testfunnorm}
\norm[\cf]{f} & \le c \sum_{k=1}^j \norm[\cf]{\hf_{n_k} u_{n_k}} = c \sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}}  \le \rho,
\\
\nonumber
\sigma_k(f) &= \begin{cases}
\displaystyle
c b^{k-j}, & k =1, \ldots, j, \\
0, & \text{otherwise},
\end{cases}
\\
\nonumber
\sigma_{k+r}(f) &= \begin{cases}
\displaystyle 
b^{r} \sigma_k(f) \le a b^r \sigma_k(f), & k+r \le j, \ r \ge 1,
\\
0 \le a b^r \sigma_k(f), & k+r > j, \ r \ge 1.
\end{cases}
\end{align}

Now suppose that $A^*(f,\varepsilon)$ samples $L_1(f), \ldots, L_n(f)$ where $n \le \comp(\rho,\varepsilon) < n_j$.  Let $u$ be a linear combination of $u_1, \cdots, u_{n_j}$ constructed to have unit norm and satisfy $L_1(u) = \cdots = L_n(u) = 0$.  Let 
\begin{gather}
\label{etadef}
f^*_\pm = f \pm \eta u, \quad \eta : = \max_{1 \le k \le j} \frac {c(a-1)b^{k-j}}{(a+1) \lambda_{n_k}}, \\
\nonumber
\norm[\cf]{f^*_\pm} \le \norm[\cf]{f} + \eta \le \frac{2ac}{a+1} \sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}}  \le \rho.
\end{gather} 
By design, $A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon)$, which will be used below.

Now we must check that $f^*_\pm \in \cc$. From the definition in \eqref{sumdef} it follows that for $k = 1, \ldots, j$ and $r \ge 1$
\begin{align*}
\sigma_k(f^*_\pm) & \begin{cases} 
\displaystyle
\le \sigma_k(f) + \eta \lambda_{n_k} = \frac{2a b^{k-j}}{a+1}, 
\\[1ex]
\displaystyle
\ge \sigma_k(f) - \eta \lambda_{n_k}  = \frac{2 b^{k-j}}{a+1}, 
\end{cases}
\\
\sigma_{k+r}(f^*_\pm) & 
\le \frac{2a b^{k+r-j}}{a+1} \le ab^r \frac{2 b^{k-j}}{a+1} \le a b^r \sigma_{k}(f^*_\pm).
\end{align*}

Given the two test functions $f^*_\pm$ and our successful algorithm, $A^*$, we note that
\begin{align*}
\varepsilon &\ge \max \bigl\{\norm[\cg]{S(f^*_+) - A^*(f^*_+,\varepsilon)}, \norm[\cg]{S(f^*_-) - A^*(f^*_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\norm[\cg]{S(f^*_+) - A^*(f,\varepsilon)} + \norm[\cg]{S(f^*_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \norm[\cg]{S(f^*_+) - S(f^*_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \norm[\cg]{S(f^*_+ - f^*_-)} \quad \text{since $S$ is linear}\\
&= \eta \norm[\cg]{S(u)} \ge \eta \lambda_{n_j} \ge \frac {c(a-1)}{(a+1)} \quad \text{by \eqref{etadef}} \\
& = \frac{\rho(a-1)}{2a} \left[\sum_{k=1}^j \frac{b^{k-j}}{\lambda_{n_{k}}} \right]^{-1}.
\end{align*}
This lower bound must be satisfied by $j$ to be consistent with the assumption $\comp(\rho,\varepsilon) < n_j$.  Any $j$ violating this inequality then provides a lower bound on $\comp(\rho,\varepsilon)$.
\end{proof}

@@

>>>>

Let $\ch$ be any subset of $\cf$, and let $\ca(\ch)$ denote the set of deterministic algorithms that successfully approximate the solution to some error tolerance for all inputs in $\ch$:
\begin{multline} \label{adapErrCrit}
\ca(\ch) : = \left\{ \text{algorithms } A:\ch \times (0,\infty) \rightarrow \cg : 
\right . \\ \left .
\bignorm[\cg]{S(f) - A(f,\varepsilon)} \le \varepsilon \ \forall f \in \ch, \ \varepsilon > 0 
\right \}.
\end{multline}
Algorithms in $\ca(\ch)$ are allowed to sample adaptively any bounded, linear functionals
of the input function $f$.  They must require only a finite number of linear functionals for each input function and positive tolerance.  The definition of $\ch$ can be used in constructing algorithms in $\ca(\ch)$, but no other a priori knowledge about the input functions is available.  Following \cite{KunEtal19a}

@@

\begin{theorem} \label{thm:lowbdcomp}
A lower bound on the complexity of the linear problem defined in \eqref{probDef} is
\begin{subequations} \label{compbd}
\begin{align}
 \label{compbdA}
\comp(&\ca(\cc),\varepsilon,\rho) \ge n_{j^*},  \\
\nonumber
\text{where} & \\
\label{compbdB}
& j^* = \max \left \{ j \in \naturals : \left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}   < 
\frac{(a-1)^2\rho^2}{(a+1)^2A^2\varepsilon^2}
\right \}.
\end{align}
\end{subequations}
\end{theorem}

@@

The first test function $f \in \cc$ is defined in terms of its series coefficients:
\begin{align}
\nonumber
\hf_i &:= \begin{cases}
\displaystyle
\frac{c b^{k-j}}{\lambda_{n_{k}}},  & i =  n_{k}, \ k = 1, \ldots, j,
\\
0, & \text{otherwise}
\end{cases}
\\
\label{cdef}
c^2 &:=  \rho^2 \left[\left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1},
\end{align}
where 
\[A = \max_{1 \le k \le j} \frac{\lambda_{n_{k-1}+1}}{\lambda_{n_k}}.\]
It can be verified that the test function lies in a ball of radius $\rho$ and also in the cone $\cc$:
\begin{align}
\label{testfunnorm}
\norm[\cf]{f}^2 &=  c^2\sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \le \rho^2,
\\
\nonumber
\sigma_k(f) &= \begin{cases}
\displaystyle
c b^{k-j}, & k =1, \ldots, j, \\
0, & \text{otherwise},
\end{cases}
\\
\nonumber
\sigma_{k+r}(f) &= \begin{cases}
\displaystyle 
b^{r} \sigma_k(f) \le a b^r \sigma_k(f), & k+r \le j, \ r \ge 1,
\\
0 \le a b^r \sigma_k(f), & k+r > j, \ r \ge 1.
\end{cases}
\end{align}

Now suppose that $A^* \in \ca(\cc)$ is an optimal algorithm, i.e., $\cost(A^*,\cc,\varepsilon,\rho) =  \comp(\ca(\cc),\varepsilon,\rho)$ for all $\varepsilon, \rho > 0$.  For our particular input $f$ defined above, suppose that $A^*(f,\varepsilon)$ samples $L_1(f), \ldots, L_n(f)$ where 
\[n + 1
\le \comp(\ca(\cc),\varepsilon,\rho) +1 < n_j.\]  Let $u$ be a linear combination of $u_1, \cdots, u_{n_j}$ constructed to have unit norm and satisfy $L_1(u) = \cdots = L_n(u) = 0$ and $\langle u,f \rangle_{\cf} = 0$. Denote
\[u = \sum_{k=1}^{j}u^{(k)}\frac{b^{k-j}}{\lambda_{n_k}}\]
where $u^{k}= \sum_{i = n_{k-1}+1}^{n_k} \hu_i u_i$ and
$\max_{1\le k\le j} \norm[\cf]{u^{(k)}} = 1.$
Let 
\begin{gather}
\label{etadef}
f^*_\pm = f \pm \eta u, \quad \eta : =  \frac {(a-1)}{(a+1)A}c, \\
\nonumber
\norm[\cf]{f^*_\pm}^2 \le \norm[\cf]{f}^2 + \norm[\cf]{\eta u } 
\le \sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \left(c^2+ \eta^2 \norm[\cf]{u^{(k)}}^2\right)\\
\nonumber
\le \sum_{k=1}^j \frac{ b^{2(k-j)}}{\lambda^2_{n_{k}}} \left(c^2+ \eta^2 \right)  \le \rho^2.
\end{gather} 
By design, $A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon)$, which will be used below.

Now we must check that $f^*_\pm \in \cc$. From the definition in \eqref{sumdef} it follows that for $k = 1, \ldots, j$ and $r \ge 1$
\begin{align*}
\sigma_k(f^*_\pm) & \begin{cases} 
\displaystyle
\le \sigma_k(f) + \sigma_k(\eta u)\le 
c b^{k-j} + \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\le b^{k-j}\left(c+\eta A \right) 
\\[1ex]
\displaystyle
\ge \sigma_k(f) - \sigma_k(\eta u)\ge 
c b^{k-j} - \eta\lambda_{n_{k-1}+1}\frac{b^{k-j}\norm[\cf]{u^{(k)} } }{\lambda_{n_k}}
\ge b^{k-j}\left(c-\eta A \right) , 
\end{cases}
\\
\sigma_{k+r}(f^*_\pm)&
\le b^{k+r-j}(c+\eta A) = ab^r b^{k-j}\frac{2c}{a+1}
=ab^r b^{k-j}\left(c-\eta A \right) \le a b^r \sigma_{k}(f^*_\pm).
\end{align*}

Given the two test functions $f^*_\pm$ and our successful algorithm, $A^*$, we note that
\begin{align*}
\varepsilon &\ge \max \bigl\{\norm[\cg]{S(f^*_+) - A^*(f^*_+,\varepsilon)}, \norm[\cg]{S(f^*_-) - A^*(f^*_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\norm[\cg]{S(f^*_+) - A^*(f,\varepsilon)} + \norm[\cg]{S(f^*_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f^*_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \norm[\cg]{S(f^*_+) - S(f^*_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \norm[\cg]{S(f^*_+ - f^*_-)} \quad \text{since $S$ is linear}\\
&= \eta \norm[\cg]{S(u)}
\end{align*}
Thus, we have
\begin{align*}
\varepsilon^2 \ge \eta \norm[\cg]{S(u)}^2 & = 
\eta^2\sum_{k=1}^{j} \norm[\cf]{S(u^{(k)}}^2 
\left(\frac{b^{k-j}}{\lambda_{n_k}}\right)^2\\
& \ge \eta^2 
\sum_{k=1}^{j} {\sigma_k(u)}^2 
\left(\frac{b^{k-j}}{\lambda_{n_k}}\right)^2\\
& \ge \eta^2 
\sum_{k=1}^{j} \norm[\cf]{u^{(k)}}^2 
b^{2(k-j)}\\
& \ge  \eta^2 b^{2(k^{*}-j)} \qquad \text{ where } k^* = \arg\max_{k} \norm[\cf]{u^{(k)}} = 1\\
& \ge \eta^2 = \frac{(a-1)^2}{(a+1)^2A^2}c^2 \\
&
=\frac{(a-1)^2 \rho^2}{(a+1)^2A^2}\left[\left(1+\frac{(a-1)^2}{(a+1)^{2} A^2 }\right)\sum_{k=1}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1}
\end{align*}

This lower bound must be satisfied by $j$ to be consistent with the assumption $\comp(\ca(\cc),\varepsilon,\rho) < n_j$.  Any $j$ violating this inequality then provides a lower bound on $\comp(\ca(\cc),\varepsilon,\rho)$.
\end{proof}

@@

\yuhannote{The figure is here in comment}
%\begin{figure}
%	\centering
%	\includegraphics[width = 8 cm]{ProgramsImages/SimDirectSolErr.eps}
%\end{figure}
